Downstream analyses of whole-genome bisulfite-sequencing data

### 2015-04-20 Rmarkdown/wgbs_downstream_analyses.Rmd
Continue revisions of WGBS downstream analyses.

\subsection{Partially methylated domains}\label{sec:pmds}

__UP TO HERE: Re-write as "Epigenome segmentation"__

Partially methylated domains, PMDs, are long stretches of the genome where the average methylation level, $B_{i}$, is away from the extremes of $0$ and $1$, typically in the range $0.20 - 0.70$. PMDs were first identified in the IMR90 genome by \citet{Lister:2009hy}. Using a simple sliding window algorithm, \citeauthor{Lister:2009hy} unexpectedly found that approximately $40\%$ of every autosome was a PMD and that the average length of these PMDs was a large 153 kb. They also showed that these PMDs were not simply due to a methylated population and an unmethylated subpopulation of cells in the sample. This did this by showing that inidividual reads mapped to these PMDs contained both methylated and unmethylated bases.

@@

PMDs have also been identified within tumour methylomes \citep{Berman:2012ga, Hansen:2011gu} and \citet{Hansen:2011gu} found that these overlap with other important genomic features called large organized chromatin lysine modifications (_LOCKs_) and lamina associated domains (_LADs_). However, PMDs are conspicuous by their absence in pluripotent cell lines, including both embryonic stem cells and induced pluripotent stem cells \citep{Lister:2011kg}. Their absence in the induced pluripotent cell lines may reflect the fact that the methylome is 'reset' upon induction of pluripotency \citep{Lister:2011kg, Stricker:2013kl}.

More sophisticated methods have also been proposed to identify these PMDs. `methylSeekR` \citep{Burger:2013kq} is one such algorithm. It uses a hidden Markov model of the $\beta$-values to segment the genome into unmethylated, lowly methylated and partially methylayed regions. While `methylSeekR` is based on $\beta$-values and does not make use of individual reads to identify PMDs, it does require the high spatial resolution of $\beta$-values that whole-genome bisulfite-sequencing provides

\section{Downstream analyses of methylation patterns at m-tuples}\label{sec:downstream_analyses_of_m-tuples}

None of the aforementioned analyses specifically exploit the fact that bisulfite-sequencing reads measure methylation patterns at multiple loci, i.e., at m-tuples. Yet much can be learnt my studying exactly that.

In this section I review the different types of questions that can be addressed when we make use of this extra information that is only available from bisulfite-sequencing data.

@@

There are several definitions and uses of _methylation entropy_ in the literature. \citet{Xie:2011cy} introduced the concept for bisulfite-sequencing data as a way of quantifying the variability of methylation patterns at an m-tuple. For a given 4-tuple, $(i, \ldots, i + 3)$, \citet{Xie:2011cy} analyse the set of reads aligned to it, $\{z: z \in \mathcal{R}_{(i, \ldots, i + 3} \}$, to compute the methylatione entropy. On the one hand, if we only observe a single unique methylation pattern then the m-tuple has the minimum methylation entropy of zero. On the other hand, if we observe all possible $2^{4} = 16$ methylation patterns then the m-tuple has the maximum methylation entropy of 1. Depending on the frequency of the patterns, we obtain intermediate methylation values ($0 < \text{methylation entropy} < 1)$. Of course, methylation entropy need not be limited to 4-tuples. The choice of the size of the tuples depends on the regions of interest, the read lengths and the sequencing coverage.

The natural interpretation of methylation entropy as a measure of 'disorder' allows us to quantify how heterogeneous DNA methylation is at a locus \citep{Xie:2011cy, He:2013cj}[^other_definitions]. These methylation entropies can be analysed to identify heterogeneous regions of the genome or perhaps tested for an association with a phenotype.

[^other_definitions]: This is closely related to the idea of identifying epialleles, for which methylation entropy has also played a role \citep{Li:2014ei} and which I discuss in Section \ref{sec:epialleles_and_epipolymorphism}. Methylation entropy has also been used to identify differential methylation, however, I do not discuss this further since this uses a different definition that is not based on analysing methylation patterns at m-tuples \citep{Zhang:2011dp, Su:2012hl}.

While \citet{Xie:2011cy} do not provide software implementing the method, a subsequent paper from some of the same researchers introduces the `DMEAS` software to do so \citep{He:2013cj} . Unfortunately, `DMEAS` is only available as a Windows binary or as a Perl script that itself is only available as PDF file, greatly limiting its usefulness.

@@

In a diploid cell, allele-specific methylation occurs when only one of the parental chromosomes is methylated at a particular locus, where the locus may be an individual cytosine or a broader region. One particular form of allele-specific methylation occurs at imprinted genes where the silenced copy of the gene is heavily methylated and the active copy is lightly methylated \citep{Shoemaker:2010jz}. However, it is now apparent that allele-specific methylation is far more prevalent than at just these imprinted regions \citep{Tycko:2010hc, Shoemaker:2010jz}.

The obvious method to detect allele-specific methylation from bisulfite-sequencing relies on the read containing a heterozygous genetic variant, such as a single nucleotide polymorphism, along with at least one methylation locus. The heterozygous variant allows the reads to be separated by the observed allele[^phase], which can then be used to test for allele-specific methylation. For example, \citet{Shoemaker:2010jz} construct a $2 \times 2$ contingency table, like that shown in Table \ref{tab:asm}, and test for an association between the allele and the methylation state using Fisher's exact test \citep{Fisher:1922bx}.

\begin{table}[h]
\centering
\begin{tabular}{@{}lll@{}}
\toprule
& Allele 1 & Allele 2 \\ \midrule
$m$ &  $m_{1}$        &  $m_{2}$       \\
$u$ &  $u_{1}$        &  $u_{2}$       \\ \bottomrule
\end{tabular}
\caption{$2 \times 2$ table used to test for allele-specific methylation.}
\label{tab:asm}
\end{table}

Such an approach, while straightforward, is also obviously limited to a subset of the available methylation loci. Both \citet{Fang:2012ci} and \citet{Peng:2012dh} published methods to detect allele-specfic methylation that do not require nearby heterozygous genetic variants nearby to the methylation locus of interest. These methods rely on the probabilistic assignment of reads to alleles, which are treated as missing data. Unfortunately, there is no publicly available software implementing the method proposed by \citet{Peng:2012dh} and so I do not discuss it further.

\citet{Fang:2012ci} uses reads containing multiple methylation loci and looks for regions of the genome where there are two distinct methylation patterns at the read-level that occur at roughly equal proportions, indicating one pattern comes from one allele and the other pattern from the other allele. The likelihood of allele-specific methylation is computed using an expectation-maximisation algorithm, which assigns reads to one of the two possible alleles. Neighbouring regions displaying allele-specific methylation are then joined together. While not mentioned in the paper, the proposed method is now available in the `MethPipe` software ([http://smithlabresearch.org/software/methpipe/](http://smithlabresearch.org/software/methpipe/)).

@@

\subsection{Epialleles and epipolymorphism}\label{sec:epialleles_and_epipolymorphism}

A DNA sequence may have multiple epigenetic states. For example, the cytosine in the sequence `TCGA` may be methylated or unmethylated; each of the methylated and unmethylated versions of that sequence is an _epiallele_. Restricting our attention to CpG methylation, a sequence with $m$ CpGs has $2^{m}$ potential epialleles. Therefore, an epiallele is just a methylation pattern at an m-tuple, with the additional constraint that the underlying DNA sequence also be identical[^epiallele_definition]. An epiallele may also be described as an _epimutation_, if it sufficiently rare.

[^epiallele_definition]: \citet{Rakyan:2002wz} impose a very strict requirement that in order for the methylation pattern to be called an epiallele it must "[result] in different phenotypes". While most would agree that is those methylation patterns that confer a difference in phenotype that are most interesting, limiting the definition of 'epiallele' to such methylation patterns seems unduly restrictive. After all, we refer to alternate forms of a genetic sequence as alleles regardless of whether we know of a phenotypic consequence of the variant.

Epipolymorphism, analogous to genetic polymorphism, is the presence of multiple epialleles in a population. However, unlike genetic polymorphism, where the population is typically a set of chromosomes from multiple individuals, the population here is often within an individual, even within a tissue within an individual. The 'epipolymorphism' of a locus has also been defined as the probability that two epialleles randomly sampled from the locus are different from one another \citep{Landan:2012kp}. \citeauthor{Landan:2012kp} define the epipolymorphism of an m-tuple as $1 - \sum_{p = 1}^{p = 2^{m}} f_{p}^{2}$, where $f_{p}$ is the estimated frequency of the $p^{th}$ methylation pattern, i.e., the number of times the $p^{th}$ pattern is observed divided by the total number of reads mapped to that m-tuple[^sampling_with_replacement].

[^sampling_with_replacement]: Strictly speaking, this is in fact an estimate of the epipolymorphism under a model that assumes sampling with replacement or, equivalently, an infinite population size. While neither assumption is true, the correction for sampling without replacement from a finite population will not substantially affect the results.

\subsubsection{Examples}

Most of the examples of epialleles with a phenotypic consequence come from the plant kingdom. Even then, the number of such epialleles is small --- a review from 2012 put the number at "about a dozen" \citep{Weigel:2012fj}.

In mammals, the study of epialleles has focused on identifying _metastable_ epialleles, which are epialleles that are mitotically heritable \citep{Rakyan:2002wz}. The poster child for the potential important of epialleles in mammals is the Agouti viable yellow ($A^{vy}$) allele \citep{Morgan:1999ds}.  Mice with an unmethylated version of the allele have a yellow coat, are obese, diabetic and have an increased suscpetibility to tumours compared to those mice with a methylated version of the allele, which have a _pseudoagouti[^pseudoagouti_vs_agouti]_ (brown) coat and none of the associated health defects.

@@

In humans, there have been several interesting studies using epialleles to infer the clonality and evolution of cancer \citep{Siegmund:2009kk, Li:2014ei}, as well as the evolution of methylation dynamics and the epipolymorphism of various loci in an immortalised cell line \citep{Landan:2012kp}.

Finally, although there is much interest, if not outright hype, surrounding transgenerational epialleles, these are difficult to prove outside of very controlled experimental conditions \citep{Daxinger:2012kl}.

Regardless of where you draw the line as to what constituted an epiallele, it has become clear that the occurrence of multiple methylation patterns at an m-tuple is the norm rather than the exception when performing bisulfite-sequencing \citep{Landan:2012kp}. There has been some progress in developing software to extract and analyse these epialleles, which I now describe.

\subsubsection{Methods}

`methclone` \citep{Li:2014ei} is software to estimate the frequency of epialleles at m-tuples and to identify "shifts" in these distributions between two samples. `methclone` is limited to pairwise comparisons. The input is a pair of `BAM` files created with the `Bismark` aligner, one for each sample. For each m-tuple (4-tuples in the example in the paper) with sufficient sequencing coverage ($60 \times$ in the example in the paper) a form of methylation entropy, called combinatorial entropy, is computed based on the set of mapped reads for the pair of samples[^methclone_output]. The "foreground" combinatorial entropy, $S$, is based on the observed frequency of epialleles in the two samples while the "background" combinatorial entropy, $\tilde{S}$, is the expected frequency of epialleles in the two samples if "all patterns of epialleles are uniformly mixed between the two stages [samples]". The difference in these combinatorial entropies, $\Delta S = S - \tilde{S}$, a kind of observed-to-expected log-ratio, is used to identify shifts in the epiallele distribution between a pair of samples. A $\Delta S = 0$ corresponds to no change and a $\Delta S = -144$ corresponds to maximal difference in entropy. It isn't clear whether the range of $\Delta S$ depends on the size of the m-tuples nor is it clear how to choose the threshold at which to declare a significant shift in the distribution of epialleles. The example in the paper shows results for a threshold of $-60, -70$ and $-80$, but these choices appear somewhat arbitrary.

[^methclone_output]: It is not clear from the paper or documentaiton whether `methclone` returns a file containing the number of methylation patterns at each m-tuple, _a ala_ `methtuple`. Unfortunately, I haven't been able to install `methclone`, so have not been able to test this out.

The obvious challenge in estimating the frequency of an epiallele is in distinguishing a _real_ epiallele from a spurious one, perhaps caused by incomplete bisulfite-conversion, sequencing error or mapping error. Another difficulty, perhaps unavoidable with the present technology, is the effect of PCR amplification bias, which will bias estimates of the relative abundance of each epiallele. `methclone` uses the observed frequencies of methylation patterns as being unbiased estimates of the true epiallele frequencies, which does not attempt to account for these potential sources of bias. In contrast, `MPFE` ([http://www.bioconductor.org/packages/release/bioc/html/MPFE.html](http://www.bioconductor.org/packages/release/bioc/html/MPFE.html), [http://f1000.com/posters/browse/summary/1097258](http://f1000.com/posters/browse/summary/1097258)), an R/Bioconductor package for "[estimating] the distribution of methylation patterns [epialleles]" at m-tuples, uses a probabilistic model to account for some of these biases.

@@

Methods designed to detect allele-specific methylation, specifically those that are based on the observed methylation patterns \citep[e.g.,][]{Fang:2012ci, Peng:2012dh}, might also adapted to identify epialleles and their associated frequencies. It is worth emphasising that since all of the methods described in this section are based entirely on the observed methylation patterns, none of these actually check that the underlying DNA sequencing is identical, which strictly speaking is a requirement for the m-tuple to be an epipolymorphic locus.

@@

A common thread in Section \ref{sec:downstream_analyses_of_m-tuples} is that while interesting methods have been proposed to address these questions, the software implementations are often lacking or, at best, very much tailored toward a particular analysis. What has been lacking is:

1. A general method for extracting methylation patterns at m-tuples,
2. Software for manipulating, analysing and visualising these methylatoin patterns.

@@

`MethylationTuples` is an R package for managing, analysing and visualising methylation patterns at m-tuples released under a Artistic-2.0 licence. R is a very popular language for analysing data, particularly in bioinformatics, and is my computational mother tongue, hence my choice of R as the language. To improve the performance of the package, parts are written in `C++`, making use of the wonderful `Rcpp` package \citep{Eddelbuettel:2011to,Eddelbuettel:2013if}.

While initially developed to support my research into co-methylation (Chapter \ref{chap:co-methylation}), the data structures developed in `MethylationTuples` are also well-suited to analysing methylation entropy, allele-specific methylation, epialleles and epipolymorphism.

`MethylationTuples` is written to work with the Bioconductor project \citep{Gentleman:2004ju}[^bioc]. Bioconductor makes extensive use of R's S4 object system and encourages developers to re-use existing Bioconductor infrastructure. From a developer's perspective, this avoids the need to re-invent the wheel. And from the user's perspective, it helps avoid multiple versions of the 'wheel', each that acts slightly differently and that may not be as well-tested.

@@

Bioconductor has excellent support for working with data defined on genomic ranges via the `IRanges` and `GenomicRanges` packages \citep{Lawrence:2013hi, Lawrence:2014gy}. However, genomic tuples, such as the co-ordinates of an m-tuples, do not naturally fit into this framework[^tuples_vs_ranges]. Therefore, I wrote a Bioconductor package for working with genomic tuples, rather unimaginatively called `GenomicTuples`, first released as part of Bioconductor version 3.0 ([http://www.bioconductor.org/packages/3.0/bioc/html/GenomicTuples.html](http://www.bioconductor.org/packages/3.0/bioc/html/GenomicTuples.html)). In fact, `GenomicTuples` is heavily based on the `GenomicRanges` package, with modifications for tuple-specific operations. For example, there is a tuple-specific method for the `findOverlaps` generic function to identify genomic tuples with equal co-ordinates (i.e., `type = \`equal'`). Since the classes in `GenomicTuples` extend those defined in `GenomicRanges`, these have excellent interoperability with existing Bioconductor infrastructure.

@@

In the `MethylationTuples` package, I define the `MethPat` class to store the genomic co-ordinates of m-tuples and the associated frequencies of the methylation patterns. The `MethPat` class extends the `SummarizedExperiment` class, also defined in the `GenomicRanges` package, but makes use of classes defined in the `GenomicTuples` package to store the genomic co-ordinates of the m-tuples. A `MethPat` object is as a matrix-like object, where rows represent m-tuples of interest and columns represent samples. Currently, it is a requirement that all m-tuples in a `MethPat` object have the same size ($\text{m}$). Figure \ref{fig:MethPat} is a schematic of a `MethPat` object storing data on $n$ samples for 3-tuples.

\begin{figure}[h]
\includegraphics[width=\textwidth]{../figures/MethPat.pdf}
\caption{Schematic of the \texttt{MethPat} class, shown here for 3-tuples. Each row represents a 3-tuple to which the genomic co-ordinates of the tuples (green box) and the frequencies of the methylation patterns (grey box) are aligned. The frequencies of each methylation pattern --- $MMM, MMU, \ldots, UUU$ --- are stored as separate matrices where the columns represent samples --- $S_{1}, \ldots, S_{n}$. Some samples may have now have any sequencing coverage for a particular m-tuple, in which case the corresponding frequencies are recorded as \texttt{NA}.}
\label{fig:MethPat}
\end{figure}

\subsection{Available methods}

The similarities to the output format of `methtuple` are clear[^methtuple_output], with the added advantage that a single `MethPat` object can contain data from multiple samples. A `MethPat` object can be constructed using the `MethPat` constructor function or from the output files of `methtuple` via the `readMethtuple` function.

@@

The `MethPat` object provides fast subsetting by rows (m-tuples), including subsetting based on overlaps of m-tuples with genomic features via the `findOverlaps` methods. Several other useful utility functions are:

- `collapseStrand`: Collapse strand-specific data by aggregating the counts. Only applicable to CpG methylation loci.
- `combine`: Combine multiple `MethPat` objects into one.
- `filterOutVariants`: Remove m-tuples that contain a known variant. Variants must be provided as a VCF.
- `getCoverage`: Compute the sequencing coverage of the m-tuples in each sample.
- `IPD`: Compute the IPD vector for each m-tuple.
- `methLevel`: Compute $\beta$-values or $\mathcal{M}$-values. Only applicable to 1-tuples.
- `MBias` and associated methods and functions: Helper functions to perform a more systematic analysis of M-bias (see Section \ref{sec:identifying_m-bias}).
- `plotBetaDensity`: Plot the kernel density of the $\beta$-values for each sample, optionally stratified by a genomic feature, e.g., CpG islands.
- `tuples`: Extract the $pos_{1}, \ldots, pos_{m}$ of the m-tuples.

These are in addition to the many useful utility functions inherited from the `SummarizedExperiment` class.

>>>>

\subsection{Epigenome segmentation}\label{sec:epigenome_segmentation}

Methylation data in the form of the vector of $\beta$-values, $\mathbf{\beta}$, may also be used by methods to segment the genome or epigenome into regulatory regions \citep{Stadler:2011kt}. One type of region that has received particular attention are so-called _partially methylated domains_ (PMDs). Partially methylated domains are long stretches of the genome where the average methylation level, $B_{i}$, is away from the extremes of $0$ and $1$, typically in the range $0.2$ to $0.7$.

PMDs were first identified in the _IMR90_ methylome by \citet{Lister:2009hy}. Using a simple sliding window algorithm, \citeauthor{Lister:2009hy} found that approximately $40\%$ of every autosome was a PMD and that the average length of these PMDs was a large 153 kb. They also showed that these PMDs were not simply due to a methylated subpopulation and an unmethylated subpopulation of cells in the sample. This did this by showing that individual reads mapped to these PMDs contained both methylated and unmethylated bases.

@@

PMDs have also been identified within tumour methylomes \citep{Berman:2012ga, Hansen:2011gu}. \citet{Hansen:2011gu} found that these PMDs overlap with other important genomic features called large organized chromatin lysine modifications (_LOCKs_) and lamina associated domains (_LADs_). However, PMDs are conspicuous by their absence in pluripotent cell lines, including both embryonic stem cells and induced pluripotent stem cells \citep{Lister:2011kg}. Their absence in the induced pluripotent cell lines may reflect the fact that the methylome is 'reset' upon induction of pluripotency \citep{Lister:2011kg, Stricker:2013kl}.

Recently, more sophisticated methods have been proposed to identify these PMDs. `methylSeekR` \citep{Burger:2013kq} is one such method. It uses a hidden Markov model of the $\beta$-values, combined with other filters, to segment the genome into unmethylated, lowly methylated and partially methylated regions.

\section{Methods based on m-tuples (m $> 2$)}\label{sec:downstream_analyses_of_m-tuples}

In this section I review the different types of questions that can be addressed by expanding our analysis to use methylation patterns at m-tuples (m $> 2$) rather than just 1-tuples. This extra information is only available from sequencing-based assays. I also describe existing software that implement some of these methods.

@@

The natural interpretation of methylation entropy is a measure of 'disorder'. It has been used to quantify how heterogeneous DNA methylation is at a locus \citep[e.g., ][]{Xie:2011cy, He:2013cj}[^other_definitions]. These methylation entropies can be analysed to identify heterogeneous regions of the genome or perhaps tested for an association with a phenotype.

[^other_definitions]: This is closely related to the idea of identifying epialleles, for which methylation entropy has also played a role \citep{Li:2014ei} and which I discuss in Section \ref{sec:epialleles}. Methylation entropy has also been used to identify differential methylation, however, I do not discuss this further since it uses a different definition that is not based on analysing methylation patterns at m-tuples \citep{Zhang:2011dp, Su:2012hl}.

Broadly speaking, on the one hand, if we only observe a single unique methylation pattern, then the m-tuple has the minimum methylation entropy of zero. On the other hand, if we observe all possible $2^{m}$ methylation patterns at equal frequency, then the m-tuple has the maximum methylation entropy (typically normalised to one). Depending on the frequency of the observed methylation patterns patterns, we obtain intermediate values of the methylation entropy.

@@

In a diploid cell, allele-specific methylation occurs when only one of the parental chromosomes is methylated at a particular locus, where the locus may be an individual cytosine or a broader region such as a gene promoter. A particularly interesting form of allele specific methylation occurs at imprinted genes, where one copy of the gene is active in a parent-specific manner. However, it is now apparent that allele-specific methylation is far more prevalent than at just these imprinted regions \citep{Tycko:2010hc, Shoemaker:2010jz}.

The obvious method to detect allele-specific methylation from bisulfite-sequencing requires reads that contaian a heterozygous genetic variant, such as a single nucleotide polymorphism, along with at least one methylation locus. The heterozygous variant allows these reads to be separated by the observed allele[^phase], which can then be used to test for allele-specific methylation. For example, \citet{Shoemaker:2010jz} construct a $2 \times 2$ contingency table, like that shown in Table \ref{tab:asm}, and test for an association between the allele and the methylation state using Fisher's exact test \citep{Fisher:1922bx}.

@@

\begin{table}[h]
\centering
\begin{tabular}{@{}lll@{}}
\toprule
& Allele 1 & Allele 2 \\ \midrule
$m$ &  $m^{1}$        &  $m^{2}$       \\
$u$ &  $u^{1}$        &  $u^{2}$       \\ \bottomrule
\end{tabular}
\caption{$2 \times 2$ table used to test for allele-specific methylation. $m^{1}$ is the number of reads with the first allele and that are also methylated at the methylation locus, $u^{1}$ is the number of reads with the first allele and that are also unmethylated at the methylation locus, etc.}
\label{tab:asm}
\end{table}

While straightforward, this approach is also obviously limited to a subset of the methylation loci in the genome. \citet{Fang:2012ci} and \citet{Peng:2012dh} published methods to detect allele-specfic methylation that do not require heterozygous genetic variants nearby to the methylation locus of interest. These methods rely on the probabilistic assignment of reads to alleles, which are treated as missing data. Unfortunately, there is no publicly available software implementing the method proposed by \citet{Peng:2012dh} and so I do not discuss it further.

\citet{Fang:2012ci} use reads containing multiple methylation loci and looks for regions of the genome where there are two distinct methylation patterns at the read-level that occur at roughly equal proportions, indicating one pattern comes from one allele and the other pattern from the other allele. The likelihood of allele-specific methylation is computed using an expectation-maximisation algorithm, which assigns reads to one of the two possible alleles. Neighbouring regions displaying allele-specific methylation are then joined together. While not mentioned in the paper, the proposed method is now available in the `MethPipe` software ([http://smithlabresearch.org/software/methpipe/](http://smithlabresearch.org/software/methpipe/)).

\subsection{Epialleles}\label{sec:epialleles}

A DNA sequence may have multiple epigenetic states. For example, the cytosine in the sequence _TCGA_ may be methylated or unmethylated; each of the methylated and unmethylated versions of that sequence is an _epiallele_. Most of the examples of epialleles with a phenotypic consequence come from the plant kingdom. Even then, the number of such epialleles is small: a review from 2012 put the number at "about a dozen" \citep{Weigel:2012fj}.

In mammals, the study of epialleles has focused on identifying _metastable_ epialleles, which are epialleles that are mitotically heritable \citep{Rakyan:2002wz}. The poster child for the potential important of epialleles in mammals is the Agouti viable yellow ($A^{vy}$) allele \citep{Morgan:1999ds}. Mice with an unmethylated version of the allele have a yellow coat, are obese, diabetic and have an increased susceptibility to tumours compared to those mice with a methylated version of the allele, which have a _pseudoagouti_[^pseudoagouti_vs_agouti] (brown) coat and none of the associated health defects.

@@

In humans, there have been several interesting studies using putative epialleles to infer the clonality and evolution of cancer \citep{Siegmund:2009kk, Li:2014ei}, as well as to study the evolution of methylation dynamics and the rate of epipolymorphism of various loci in an immortalised cell line \citep{Landan:2012kp}.

Regardless of where you draw the line as to what constitutes an epiallele, it has become clear in the analysis of bisulfite-sequencing data that the occurrence of multiple methylation patterns at an m-tuple is the norm rather than the exception.

Restricting our attention to CpG methylation, a sequence with $m$ CpGs has $2^{m}$ potential epialleles. In other words, an epiallele is just a methylation pattern at an m-tuple, with the additional constraint that the underlying DNA sequence also be identical[^epiallele_definition]. An epiallele may also be described as an _epimutation_ if it is different from the 'normal' methylation state.

[^epiallele_definition]: \citet{Rakyan:2002wz} impose a very strict requirement that in order for the methylation pattern to be called an epiallele it must "[result] in different phenotypes". While most would agree that it is those methylation patterns that confer a difference in phenotype that are most interesting, limiting the definition of 'epiallele' to such methylation patterns seems unduly restrictive. After all, we refer to alternate forms of a genetic sequence as alleles regardless of whether we know of a phenotypic consequence of the variant.

The rate of _epipolymorphism_ of a locus is defined as the probability that two epialleles randomly sampled from the locus are different from one another \citep{Landan:2012kp}[^epipolymorphism]. \citeauthor{Landan:2012kp} define the rate of epipolymorphism of an m-tuple as $1 - \sum_{p = 1}^{p = 2^{m}} f_{p}^{2}$, where $f_{p}$ is the estimated frequency of the $p^{th}$ methylation pattern, i.e., the number of times the $p^{th}$ pattern is observed divided by the total number of reads mapped to that m-tuple[^sampling_with_replacement]. Unlike genetic polymorphisms, where the population is typically a set of chromosomes from multiple individuals, the population of epipolymorphisms is often within an individual, even within a tissue within an individual.

[^epipolymorphism]: \citet{Landan:2012kp} actually call this the "epipolymorphism" of the locus rather than the "rate of epipolymorphism" of the locus. However, I think this is better described as a rate since it refers to the frequency at which we observe epialleles/epipolymorphisms.

[^sampling_with_replacement]: Strictly speaking, this is in fact an estimate of the epipolymorphism of the locus under a model that assumes sampling with replacement or, equivalently, an infinite population size. While neither assumption is true, the correction for sampling without replacement from a finite population will not substantially affect the results.

The obvious challenge in estimating the frequency of an epiallele is in distinguishing a _real_ epiallele from a spurious one, perhaps caused by incomplete bisulfite-conversion, sequencing error or mapping error. Another difficulty, perhaps unavoidable with current technology, is the effect of PCR amplification bias, which will bias estimates of the relative abundance of each epiallele.

Of the downstream analyses based on methylation patterns at m-tuples, the study of epialleles has received the most attention with respect to methods and software.

`methclone` \citep{Li:2014ei} is a method to estimate the frequency of epialleles at m-tuples (the rate of epipolymorphism) and to identify "shifts" in these distributions between a pair of samples. `methclone` is based on computing and comparing two forms of methylation entropy, the "foreground" and "background". The foreground combinatorial entropy, $S$, is based on the observed frequency of epialleles in the two samples. The background combinatorial entropy, $\tilde{S}$, is the expected frequency of epialleles in the two samples if "all patterns of epialleles are uniformly mixed between the two [samples]". The difference in these combinatorial entropies, $\Delta S = S - \tilde{S}$, a kind of observed-to-expected log-ratio, is used to identify shifts in the epiallele distribution between a pair of samples. A $\Delta S = 0$ corresponds to no change and a $\Delta S = -144$ corresponds to maximal difference in entropy. It isn't clear whether the range of $\Delta S$ depends on the size of the m-tuples nor is it clear how to choose the threshold at which to declare a significant shift in the distribution of epialleles.

`methclone` uses the observed frequencies of methylation patterns as being unbiased estimates of the true epiallele frequencies and does not attempt to account for potential sources of bias. In contrast, `MPFE` ([http://www.bioconductor.org/packages/release/bioc/html/MPFE.html](http://www.bioconductor.org/packages/release/bioc/html/MPFE.html), [http://f1000.com/posters/browse/summary/1097258](http://f1000.com/posters/browse/summary/1097258)), an R/Bioconductor package for "[estimating] the distribution of methylation patterns [epialleles]" at m-tuples, uses a probabilistic model to account for some of these biases.

@@

Methods designed to detect allele-specific methylation, specifically those that are based on the observed methylation patterns \citep[e.g.,][]{Fang:2012ci, Peng:2012dh}, might also adapted to identify epialleles and their associated frequencies. It is worth emphasising that since all of the methods described in this section are based entirely on the observed methylation patterns, none of these actually check that the underlying DNA sequencing is identical, which, strictly speaking, is a requirement for the m-tuple to be an epipolymorphic locus.

\subsection{Software for analysing methylation patterns at m-tuples}\label{sec:methylation_pattern_software}

Generally speaking, there are fewer software options for analysing methylation patterns at m-tuples (m $> 2$) than there are for analysing methylation patterns at 1-tuples. Furthermore, the available options are often difficult to extend since they are typically developed for a specific task and not for general computations with methylation patterns at m-tuples.

I have also experienced considerable difficulty in applying some of these methods owing to poor software implementations. Two give two examples, `DMEAS` \citep{He:2013cj is only available as a Windows binary or as a Perl script that itself is only available as a PDF file and I have been unable to install `methclone` __TODO: Give reasons__.

`MethPipe` is perhaps the best documented and potentially extensible software for analysing methylation patterns at m-tuples. `MethPipe` is mostly written in C++ and is designed as a suite of tools for a complete 'pipeline' analysis of bisulfite-sequencing data. As such, it does not feature tools that are particularly amenable to interactive or exploratory analyses. In fact, I am unaware of any software that allows easy exploratory analyses of methylation patterns at m-tuples, which in part motivated the development of `MethylationTuples`.

@@

In order to facilitate the development of downstream analysis methods based on methylation patterns at m-tuples, I saw the need for two pieces of software:

1. Software for extracting methylation patterns at m-tuples,
2. Software for manipulating, analysing and visualising these methylation patterns.

@@

`MethylationTuples` is an R package for managing, analysing and visualising methylation patterns at m-tuples, It is released under an Artistic-2.0 licence (__TODO: Why use this instead of MIT/GNU?__). I chose to write this software in R because it is a very popular language for data analysis, particularly in bioinformatics, and facilitates both batch and interactive usage. R is also my computational mother tongue and an R package is a convenient unit for sharing reusable code. To improve the performance of key functionality, parts of `MethylationTuples` are written in `C++`, making use of the wonderful `Rcpp` package \citep{Eddelbuettel:2011to,Eddelbuettel:2013if}.

While initially developed to support my research into co-methylation (Chapter \ref{chap:co-methylation}), the data structures developed in `MethylationTuples` are well-suited to any analysis based on methylation patterns of m-tuples such as methylation entropy, allele-specific methylation and the identification of epialleles. Of course, `MethylationTuples` can equally be used to develop methods based on 1-tuples, such as identifying differential methylation, since 1-tuples are just a particular type of m-tuple.

`MethylationTuples` is written to work within the Bioconductor project \citep{Gentleman:2004ju}[^bioc]. Bioconductor makes extensive use of R's S4 object system and encourages developers to re-use existing Bioconductor infrastructure. From a developer's perspective, this avoids the need to re-invent the wheel when tackling common tasks. And from the user's perspective, it helps avoid multiple versions of the 'wheel', each that might otherwise act slightly differently and that may not be as well-tested.

@@

Bioconductor has excellent support for working with data defined on genomic ranges via the `IRanges` and `GenomicRanges` packages \citep{Lawrence:2013hi, Lawrence:2014gy}. However, genomic tuples, such as the co-ordinates of an m-tuples, do not naturally fit into this framework[^tuples_vs_ranges]. Therefore, I first wrote a Bioconductor package for working with genomic tuples, rather unimaginatively called `GenomicTuples`, first released as part of Bioconductor version 3.0 ([http://www.bioconductor.org/packages/3.0/bioc/html/GenomicTuples.html](http://www.bioconductor.org/packages/3.0/bioc/html/GenomicTuples.html)). In fact, `GenomicTuples` is heavily based on the `GenomicRanges` package, with modifications for tuple-specific operations. For example, there is a tuple-specific method for the `findOverlaps` generic function to identify genomic tuples with equal co-ordinates (i.e., \code{type = `equal'}). Since the classes in `GenomicTuples` extend those defined in `GenomicRanges`, these have excellent interoperability with existing Bioconductor infrastructure.

@@

In the `MethylationTuples` package, I define the `MethPat` class to store the genomic co-ordinates of m-tuples and the associated counts of each methylation pattern. A `MethPat` object is as a matrix-like object, where rows represent m-tuples of interest and columns represent samples. The `MethPat` class extends the `GenomicRanges::SummarizedExperiment`[^namespace] class but makes use of classes defined in the `GenomicTuples` package to store the genomic co-ordinates of the m-tuples. Currently, it is a requirement that all m-tuples in a `MethPat` object have the same size (i.e., same $\text{m}$).

[^namespace]: This uses the NAMESPACE notation of R: `GenomicRanges::SummarizedExperiment` can be read as "the `SummarizedExperiment` class is part of the `GenomicRanges` package".

Figure \ref{fig:MethPat} is a schematic of a `MethPat` object storing methylation patterns at 3-tuples for $n$ samples. The similarities to the output format of `methtuple` are clear[^methtuple_output], with the added advantage that a single `MethPat` object can contain data from multiple samples.

\begin{figure}[h]
\includegraphics[width=\textwidth]{../figures/MethPat.pdf}
\caption{Schematic of the \texttt{MethPat} class, shown here for 3-tuples. Each row represents a 3-tuple to which the genomic co-ordinates of the tuples (green box) and the counts of the methylation patterns (grey box) are aligned. The counts of each methylation pattern ($MMM, MMU, \ldots, UUU$) are stored as separate matrices where the columns represent samples ($S_{1}, \ldots, S_{n}$). Some samples may not have any sequencing coverage for a particular m-tuple, in which case the corresponding frequencies are recorded as \texttt{NA}.}
\label{fig:MethPat}
\end{figure}

@@

\subsection{Methods}

A `MethPat` object can be constructed directly using the `MethPat()` constructor function or from the output files of `methtuple` via the `readMethtuple()` function.

The `MethPat` object provides fast subsetting by rows (m-tuples) and columns (samples) via the `[` method. It also benefits from fast subsetting based on overlaps of m-tuples with genomic features via the `findOverlaps()`-based methods. Several other useful utility functions are:

__UP TO HERE: Check these exist, update descriptions and add any new functions__
